
\section*{Acknowledgement}
\addcontentsline{toc}{section}{Acknowledgement \vspace{-0.3cm}}

Ostrouchov, Schmidt, and Patel were supported in part by the project
``NICS Remote Data Analysis and Visualization Center''
funded by the Office of Cyberinfrastructure of the
U.S. National Science Foundation
under Award No. ARRA-NSF-OCI-0906324 for NICS-RDAV center.
Chen and Ostrouchov were supported in part by the project
``Visual Data Exploration and Analysis of Ultra-large Climate Data''
funded by U.S. DOE Office of Science
under Contract No. DE-AC05-00OR22725.
\\

This work used resources of National Institute for Computational
Sciences at the University of Tennessee, Knoxville, which is supported
by the Office of Cyberinfrastructure of the U.S. National Science Foundation
under Award No. ARRA-NSF-OCI-0906324 for NICS-RDAV center.
This work also used resources of the Oak Ridge Leadership Computing Facility
at the Oak Ridge National Laboratory, which is supported by the Office of
Science of the U.S. Department of Energy under Contract No. DE-AC05-00OR22725.
This work used resources of the Newton HPC Program at the University of Tennessee, 
Knoxville. 
\\

We thank our colleague, Ed D'Azevedo
from the Computational Mathematics Group,
Computer Science and Mathematics Division,
Oak Ridge National Laboratory (ORNL),
for his discussions and illuminating advice using 
ScaLAPACK and distributed matrix computation.
\\

We also thank Brian D. Ripley, Kurt Hornik, Uwe Ligges, and Simon Urbanek
from the \proglang{R} Core Team for discussing package release issues and
helping us solve portability problems on different platforms.
\\
